### 2025-06-24 thesis_julio/Chapter1/chapter1.tex
thesis24june

\begin{figure}[h]
    \centering
   \includegraphics[width=1\textwidth]{sm.png}
    \caption{The standard model of particle physics}
    \label{fig:four-forces}
\end{figure}

The fermionic sector contains particle the particles that make up all known matter in the universe and is divided into the leptons and the quarks, both of these particles are divided into three generations, each generation being heavier than the one before. For the leptons we have the electron and Its neutrino for the first generation, the second generation is. the $\mu$ and Its neutrino and the third one is the $\tao$ with Its neutrino. In the case of quarks we have the quark up and the quark down for the first generation, the second generation que have the quark charmd and the quark strange and for the third generation que have the quark top and the quark bottom. The bosons are the mediator particles called the gauge bosons already mentions, the photon, the gluon, the bosons $W\pm$ and the boson Z. There is also de Higgs bosson, which is the last gauge bossons and It's the responsible to give mass to the other SM particles

One of the main sources to obtain elementary particles is particle accelerators, in this you accelerate a particle into high energy and smash them with a target, with the proper  arrangements of magnets you can study the debris from the collision, for more heavy particles you need higher level of energy to the collision.

@@

The Large Hadron Collider (LHC) is the biggest and powerful particle accelerator in the world at this moment, It is a 27 kilometers ring of superconducting magnets, this magnets boost the particles energies to obtain speeds that are close to the light speed before collisions. These particles are introduced as to the accelerator as particles beams this beams travel in opposing directions in tubes that are kept at ultrahigh vacuum, with the help of a electromagnetic field made by the electromagnets operating in superconduct state at low temperatures thanks to liquid helium this particles are guided across the whole accelerator. \cite{LHC}

\begin{figure}[h]
    \centering
   \includegraphics[width=1\textwidth]{LHC.png}
    \caption{The accelerator facility at cern for the LHC}
    \label{fig:LHC}
\end{figure}

The LHC is part of an accelerator complex in CERN that is a succession of different machines that increase the energy of the beam of particles before passing into the next one on the sequence the particles at last are introduced on the LHC which is the last element in the sequence. The particles start at the linear accelerator 4 (Linac4) which is the source of the protons beams, it accelerates the particles in this case negative hydrogen ions to 160 MeV to enter the Proton Synchroton Booster (PSB), this is were the hydrogen loses Its two electrons leaving only protons and accelerating the beam to 2 GeV  and then are injected into the Proton Synchroton (PS), in where the protons are accelerated up to 26 GeV and then they are sent into the Super Proton Synchroton (SPS) here they are accelerated up to 450 GeV and then they're finally introduced into two beams pipes on the LHC, it takes about 4 minutes and 20 seconds to fill the LHC rings reaching energies up to 6.5 TeV, there are 4 collisions points where the detectors are located Alice, Atlas, CMS and LHCb, in the collision the total energy is of 13.6 TeV. \cite{LHCII}

The LHC has been working since 2009 with the discovery of the Higgs Boson on 2012 and looking for new physics It has been working on different periods called Runs, Run 1 was on the period of (2009-2013), Run 2 was on (2015-2018), Run 3 is currently on going since 2022.

@@

The performance of a collider is determined by the beam energy and the luminosity. First we are going to define the cross section in a classical way the cross section area is parameter of interest that could say that measure the size of a target, this cross sectional area is presented to a stream of incoming particles, this is similar in the case of the scattering in elementary particles, if you fire a electron beam into a proton tank the parameters of interest are the size of the proton and the cross sectional area $\sigma$ that presents the incident beam. This collision differs to the classic case in some ways, first you could call the target soft since It's not a possible collision more like the closer you get the greater the deflection, the electrons scatters off the hydrogen, various interactions are happening during this and each one has a It's own scattering cross section $\sigma_{i}$ nevertheless in the experiment we are interested in the total cross section which can be seen as the sum of all the cross sections.

@@

Each of this cross sections depends typically in the velocity of the incident particle, It's expected that the cross section would be proportional to the time that the incident particle spends close to It's target which means that $\sigma$ is inversely proportional to the velocity but this behavior is altered close to the region of a energy that particles involved to interact forming a short lived semi-state before breaking apart, things like this make the cross section for elementary particles different to the classic one.

To explain the cross section when a target is soft lets supposed a electron encounter a potential and scatters at the angle $\theta$, this scatter is angle is a function of a impact parameter b the distance by which this particle would have missed the scattering center had it continued on its original trajectory. The form of $\theta(B)$ depends on the potential and the smaller the impact parameter b, larger the deflection. If the particle comes with an impact parameter between b and b + db it will merge in a angle between $\theta$ and $\theta + d\theta$ and even in a more general case if it passes through a infinitesimal area $d\sigma$ it will scatter into a corresponding solid angle $d\Omega$, the larger $d\sigma$ is the larger $d\Sigma$ the proportionality factor between these two is called the differential cross section D this can be illustrated in the next image:

@@

This is a classical description of scattering and could be seen as the one who controls the number of particles going through the detector also determined the solid angle. Here we have the incident flux F and a cross sectional area and are a way to interpret collision rates in modern particle accelerators since in high energy physics this structure is similar with a quantity called luminosity L with Its own cross section $\sigma$.

Even when the dynamics are different for the quantum field theory case, the general principle remains the same, when we have a  higher the cross section we obtain a higher luminosity and Its more likely to observer determined events.

%The define of cross section can be understand in a situation where a single particle is travelling to a velocity $v _{a}$in a region defined by A, which contains $n_{b}$ particles of a particle type b moving to a velocity $v_{b}$ in the opposite direction of $v_{a}$ if in a time $\delta t$  the particle a cross a region contained a total of $nb(v_{a} + v_{b})A\delta t$ particles of type b, the interaction probability can be obtained from the effective total cross seciotanl area divided by A which can be seen as the probability of that hte incident particles passes through one of the region of area $\sigma$ around each of the $\delta N$ particles then we can talk about the total rate being:

@@

In which R is the number of events per second or rate, the $\sigma_{p}$ is the cross section which in the quantum case is a probability of interaction which can be obtained from the standar model of particle physics and L the instantanious luminosity. The unity of L is  $cm^{-2} s^{-1}$ a higher luminosity means greater probability that the particles will collide and produce the desired interactions, luminosity can be increased then in two ways, packing more particles into the beams or focusing this beams more.

To compute the luminosity of two there is a few things that need to be taken in consideration. First the density distribution of each beam in the transverse and longitudinal plane, with the. two beams moving towards each other, the position and the time as the cross each other also need to be considered and calling the distance to the collision point $s_{0}$. The figure (1.x) shows two beams colliding which depends on a variety of factors:

@@

\begin{equation}
 L = 2N_{1}N{2}fN_{b} \int \int \int^{\infty}_{-\infty}  \rho_{1x} (x) \rho_{1y}(y) \rho_{1s} (s-s_{0}) \rho_{2x}(x) \rho_{2y}(y) \rho_{2s} (s+s_{0}) dxdydsds_{0}  
\end{equation}

@@

This is related directly to the number of observed events $L_{int} \cdot \sigma_{p}$ number of events of interest.

@@

The PCC luminometer method consist in counting the average number of pixel clusters on the detector to measure luminosity. This occurs during a zero bias event, an event that is triggered by requiring only that two proton bunches cross at the CMS interaction point. Giving that the number of pixels in really big the probability that a pixel is being hit by two different tracks by the same bunch crossing is really small. The mean number of pixel clusters in a simulated zero bias event is in the order of 100 per pp collision. The number of pixel clusters per bunch crossing is linearly dependent on pileup and therefore an accurate measure of the instantaneous luminosity.\cite{PCC1} The mean number of pixel cluster per trigger is

\begin{equation}
<N_{cluster}> = <N_{Pixel/Interaction}> <N_{interaction}> = <N_{Pixel/interaction}> \mu 
\end{equation}

here the $\mu$ is the number of interactions and with this the following equation and gives us a relation with the cross section $\simga_{i}$, instant luminosity and $\mu$ and for the obtention of the luminosity value:

\begin{equation}
<N_{cluster}> = \frac{\sigma_{cluster}}{f} \frac{dL}{dt}
\end{equation}

@@

There have been different luminosities reached on the LHC for the Run 1 a luminosity of $0.77 x 10^{34}$ was reached and a integrated luminosity of 25 $fb^{-1}$  with a precision of 2.0\% for the first part and second part of the Run 2 the luminosity measured was of 38.4 $fb^{-1}$ with a precision of 1.3\% and 78$fb^{-1}$ respectively\cite{LHClum}.  The following image show shows the integrated luminosity for the CMS experiment on different years.

\begin{figure}[h]
    \centering
     \includegraphics[scale=1.25]{integratedlum.png}
     \caption{Integrated luminosity at CMS}
    \label{fig:CMS-Luminosity}
\end{figure}

>>>>

\begin{figure}[h]
    \centering
   \includegraphics[width=1\textwidth]{sm.png}
    \caption{The standard model of particle physics.\cite{pic2}}
    \label{fig:four-forces}
\end{figure}

The fermionic sector contains particle the particles that make up all known matter in the universe and is divided into the leptons and the quarks, both of these particles are divided into three generations, each generation being heavier than the one before. For the leptons we have the electron and It's neutrino for the first generation, the second generation is the $\mu$ and It's neutrino and the third one is the $\tau$ with It's neutrino. In the case of quarks we have the quark up and the quark down for the first generation, the second generation we have the quark charm and the quark strange and for the third generation we have the quark top and the quark bottom. The bosons are the mediator particles, the photon, the gluon, the bosons W+, W- and the boson Z. There is also de Higgs boson which is the responsible to give mass to the other SM particles

One of the main sources to obtain elementary particles is particle accelerators, in this you accelerate a particle into high energy and smash them with a target, with the proper  arrangements of magnets you can study the debris from the collision, to produce heavy particles you need higher level of energy to the collision.

@@

The Large Hadron Collider (LHC) is the biggest and powerful particle accelerator in the world at this moment. It is a 27 kilometers ring of superconducting magnets, this magnets mantain the particle in orbits meanwhile meanwhile radio frequency cavities increase the particles energies to obtain speeds that are close to the light speed before collisions. These particles are introduced into the accelerator as beams that are traveling in opposing directions in tubes that are kept at ultrahigh vacuum. All of this with the help of a electromagnetic field made by the electromagnets operating in superconduct state at low temperatures thanks to liquid helium this particles are guided across the whole accelerator. \cite{LHC}

\begin{figure}[h]
    \centering
   \includegraphics[width=1\textwidth]{LHC.png}
    \caption{The accelerator facility at cern for the LHC. \cite{lhcpic}}
    \label{fig:LHC}
\end{figure}

The LHC is part of an accelerator complex in CERN that is a succession of different machines that increase the energy of the beam of particles before passing into the next one on the sequence the particles at last are introduced on the LHC which is the last element in the sequence. The particles start at the linear accelerator 4 (Linac4) which is the source of the protons beams, hydrogen ions to 160 MeV to enter the Proton Synchroton Booster (PSB), this is were the hydrogen loses Its two electrons leaving only protons and accelerating the beam to 2 GeV  and then are injected into the Proton Synchroton (PS). Here the protons are accelerated up to 26 GeV and then they are sent into the Super Proton Synchroton (SPS) here they are accelerated up to 450 GeV. Finally they're fintroduced into two beams pipes on the LHC, it takes about 4 minutes and 20 seconds to fill the LHC rings reaching energies up to 6.5 TeV. There is a total of four collisions points and this is where the detectors are located Alice, Atlas, CMS and LHCb, in the collision the total energy is of 13.6 TeV. \cite{LHCII}

The LHC has been working since 2009 with the discovery of the Higgs Boson on 2012 and looking for new physics. It has been working on different periods called Runs, Run 1 was on the period of (2009-2013), Run 2 was on (2015-2018). Run 3 is currently on going since 2022.

@@

The performance of a collider is determined by the beam energy and the luminosity. First we are going to define the cross section in a classical way, the cross section area is parameter of interest that measures the size of a target, this cross sectional area is presented to a stream of incoming particles, similar to the case of the scattering in elementary particles, if you fire a electron beam into a proton, the parameters of interest are the size of the proton and the cross sectional area $\sigma$ that presents the incident beam.

This collision differs to the classic case in some ways, first you could call the proton target soft since It's not a possible collision, instead as the electron becomes closer it gradually deflects due to the electromagnetic interactions. There are two main interactions each that can happen in this case, the first one the elastic scattering, here neither the electron or the proton changes their internal states. The other one the inelastic scattering where the target absorbs energy. In the case of the later the cross section $\sigma_{total}$ would be the sum of all the possible inelastic process $\sigma_{i}$ as shown in equation (1.1).

@@

Each of this cross sections depends typically in the velocity of the incident particle, It's expected that the cross section would be proportional to the time that the incident particle spends close to Ithe target which means that $\sigma$ is inversely proportional to the velocity. This behavior is altered close to the region of energy that particles involved to interact forming a short lived semi-state before breaking apart, things like this make the cross section for elementary particles different to the classic one.

To understand an elastic process lets consider an electron beam which encounters a potential and scatters at the angle $\theta$, this scatter is angle is a function of a impact parameter b the distance by which this particle would have missed the scattering center had it continued on its original trajectory. The form of $\theta(B)$ depends on the potential and the smaller the impact parameter b, larger the deflection. If the particle comes with an impact parameter between b and b + db it will merge in a angle between $\theta$ and $\theta + d\theta$ and even in a more general case if it passes through a infinitesimal area $d\sigma$ it will scatter into a corresponding solid angle $d\Omega$, the larger $d\sigma$ is the larger $d\Sigma$ the proportionality factor between these two is called the differential cross section D this can be illustrated in the next image:

@@

This is a classical description of scattering and could be seen as the one who controls the number of particles going through the detector also determined the solid angle. 
%The define of cross section can be understand in a situation where a single particle is travelling to a velocity $v _{a}$in a region defined by A, which contains $n_{b}$ particles of a particle type b moving to a velocity $v_{b}$ in the opposite direction of $v_{a}$ if in a time $\delta t$  the particle a cross a region contained a total of $nb(v_{a} + v_{b})A\delta t$ particles of type b, the interaction probability can be obtained from the effective total cross seciotanl area divided by A which can be seen as the probability of that hte incident particles passes through one of the region of area $\sigma$ around each of the $\delta N$ particles then we can talk about the total rate being:

@@

In which R is the number of colision events per second or rate, the $\sigma_{p}$ is the cross section which in the quantum case is a probability of interaction which can be obtained from the standar model of particle physics and L the instantanious luminosity. The unity of L is  $cm^{-2} s^{-1}$ a higher luminosity means greater probability that the particles will collide and produce the desired interactions, luminosity can be increased then in two ways, packing more particles into the beams or focusing this beams more.

To compute the luminosity there is a few things that need to be taken in consideration. First the density distribution of each beam in the transverse and longitudinal plane, with the two beams moving towards each other, the position and the time as the cross each other also need to be considered and calling the distance to the collision point $s_{0}$. The figure (1.5) shows two beams colliding which depends on a variety of factors:

@@

\begin{equation}
 L = 2N_{1}N_{2}fN_{b} \int \int \int \int^{\infty}_{-\infty}  \rho_{1x} (x) \rho_{1y}(y) \rho_{1s} (s-s_{0}) \rho_{2x}(x) \rho_{2y}(y) \rho_{2s} (s+s_{0}) dxdydsds_{0}  
\end{equation}

@@

This is related directly to the number of observed events $L_{int} \cdot \sigma_{p}$.

@@

The PCC luminometer method consist in counting the average number of pixel clusters on the detector to measure luminosity. This occurs during a zero bias event, an event that is triggered by requiring only that two proton bunches cross at the CMS interaction point. Giving that the number of pixels is really big the probability that a pixel is being hit by two different tracks by the same bunch crossing is really small. The mean number of pixel clusters in a simulated zero bias event is in the order of 100 per pp collision. The number of pixel clusters per bunch crossing is linearly dependent on pileup and therefore an accurate measure of the instantaneous luminosity.\cite{PCC1} The mean number of pixel cluster per trigger is

\begin{equation}
<N_{cluster}> = <N_{Cluster/Interaction}> <N_{interaction}> = <N_{Cluster/interaction}> \mu 
\end{equation}

here the $\mu$ is the number of interactions, now we define the collision rate in the equation (1.8):

\begin{equation}
R = \mu * f_{LHC} = L * \sigma_{p}
\end{equation}

Solving for $\mu$ we got:

\begin{equation}
\mu = \frac{L * \sigma_{p}}{f_{LHC}} 
\end{equation}

And combining equation (1.7) and (1.9) we obtain:

\begin{equation}
<N_{cluster}> = \frac{\sigma_{p}* L}{f_{LHC}}  <N_{Cluster/Interaction}>
\end{equation}

@@

There have been different luminosities reached on the LHC for the Run 1 a luminosity of $0.77 x 10^{34} cm^{-2} s^{-1}$ was reached and a integrated luminosity of 25 $fb^{-1}$  with a precision of 2.0\% for the first part and second part of the Run 2 the luminosity measured was of 38.4 $fb^{-1}$ with a precision of 1.3\% and 78$fb^{-1}$ respectively\cite{LHClum}.  The figure (1.6) show shows the integrated luminosity for the CMS experiment on different years.

\begin{figure}[h]
    \centering
     \includegraphics[scale=1.25]{integratedlum.png}
     \caption{Integrated luminosity at CMS.}
    \label{fig:CMS-Luminosity}
\end{figure}
